Contextuality and Nonlocality in Discrete Quantum Theory

### 2016-01-18 DQC3.tex
Update "6.1 Gleason's Theorem in DQT"...

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% User specified LaTeX commands.
\usepackage{amsfonts}
\usepackage{color}
\usepackage{stmaryrd}\usepackage{dsfont}
\usepackage[usenames,dvipsnames]{xcolor}
\usepackage{enumitem}		% customizable list environments
\newlength{\lyxlabelwidth}      % auxiliary length 
\theoremstyle{plain}
\newtheorem{thm}{\protect\theoremname}
\ifx\proof\undefined
\newenvironment{proof}[1][\protect\proofname]{\par
\normalfont\topsep6\p@\@plus6\p@\relax
\trivlist
\itemindent\parindent
\item[\hskip\labelsep\scshape #1]\ignorespaces
}{%
\endtrivlist\@endpefalse
}
\providecommand{\proofname}{Proof}
\fi
\theoremstyle{definition}
\newtheorem{problem}[thm]{\protect\problemname}

@@

%%%%%%%%%%%%%%%%%
\subsection{A No-Go Theorem}

@@

\yutsung{I am going to prove a no-go theorem for $d=3$ here with
an extra property
\begin{description}
\item [{(iv)}] $\dpt(\ket{\Phi},\ket{\Psi})=\dpt(\lambda\ket{\Phi},\ket{\Psi})=\dpt(\ket{\Phi},\lambda\ket{\Psi})$
where $\lambda\in\ff{p^{2}}^{*}$. 
\end{description}
This property means that the phase difference and normalization will
not effect the probability assignment. Although this property will
not be used explicitly, considering non-normalized states may simplify
the proof. 
\begin{thm}
\label{theorem1-1} There is no map $\dpt\colon\ffzd{p^{2}}\times\ffzd{p^{2}}\longrightarrow[0,1]$
satisfying properties ({\bf i}), ({\bf ii}), ({\bf iii}),
and ({\bf iv}) when $d\ge3$ and $p>3$.\end{thm}
}

@@

%%%%%%%%%%%%%%%%%
\subsection{Interval-Valued Probabilities for DQT}

@@

Recall the measurement process discussed in section \ref{sec:Critical-Analysis-CQT}.
Given a physical system represented by a state $\ket{\Psi}$, and
an observable ${\mathbf{O}}$, whose eigenspectrum is given by 
\[
{\mathbf{O}}\left|i\right\rangle =\lambda_{i}\left|i\right\rangle \ .
\]
Then the Born rule of CQT tells us that the probability of obtaining
the eigenvalue $\lambda_{i}$ as an outcome of the measurement is
given by 
\begin{eqnarray}
\begin{array}{rcccl}
{\cal P}_{\Psi}\left(\ket{i}\right)={\cal P}_{\Psi}\left(\lambda_{i}\right) & = & {\displaystyle {\frac{\langle\Psi|{P}_{i}|\Psi\rangle}{\ip{\Psi}{\Psi}}}} & = & {\displaystyle {\frac{\ip{\Psi}{i}\ip{i}{\Psi}}{\ip{\Psi}{\Psi}\ip{i}{i}}}}=\Tr\left(\frac{\proj{i}}{\ip{i}{i}}\frac{\proj{\Psi}}{\ip{\Psi}{\Psi}}\right)\end{array}\ ,\label{prob1.eq-6}
\end{eqnarray}
If the system is a mixed state represented by a density matrix 
\[
\rho=\sum_{j}p_{j}\frac{\proj{\Psi_{j}}}{\ip{\Psi_{j}}{\Psi_{j}}}\textrm{ ,}
\]
equation (\ref{prob1.eq-6}) can be extended to 
\begin{eqnarray}
{\cal P}_{\rho}\left(\ket{i}\right)={\cal P}_{\rho}\left(\lambda_{i}\right) & = & \Tr\left(\frac{\proj{i}}{\ip{i}{i}}\sum_{j}p_{j}\frac{\proj{\Psi_{j}}}{\ip{\Psi_{j}}{\Psi_{j}}}\right)=\sum_{i}p_{i}\frac{\ip{\Psi_{j}}{i}\ip{i}{\Psi_{j}}}{\ip{\Psi_{j}}{\Psi_{j}}\ip{i}{i}}\label{prob1.eq-2}
\end{eqnarray}
Let $\C^{d*}=\left\{ \ket{\Psi}\in\C^{d}\mid\ip{\Psi}{\Psi}\ne0\right\} $,
then ${\cal P}_{\rho}:\C^{d*}\rightarrow\left[0,1\right]$ defines
a probability which means ${\cal P}_{\rho}$ satisfies 
\begin{description}
\item [{(i')}] For any orthonormal basis $\{\ket{i}\}$ and any state $\ket{\Psi}$,
${\displaystyle \sum_{i=0}^{d-1}{\cal P}_{\rho}\left(\ket{i}\right)=1}$. 
\end{description}
Conversely, if an assignment ${\cal P}:\C^{d*}\rightarrow\left[0,1\right]$
is a probability, i.e., satisfying ({\bf i'}), we can identify
${\cal P}$ to a mixed state $\rho$ by Gleason's theorem~\cite{Redhead1987-REDINA,gleason1957}\footnote{In order to simplify the statement, we consider consider the probability
on the state vectors instead of the projectors or closed subspaces
as in the Gleason's original paper.}.
\begin{thm}
(Gleason's Theorem) For $d\ge3$, for any probability ${\cal P}:\C^{d*}\rightarrow\left[0,1\right]$,
there is a density matrix $\rho$ such that ${\cal P}={\cal P}_{\rho}$.
\end{thm}
Consequently, the Born rule is the only reasonable probability assignment,
and can be deduced by other reasonable axioms~\cite{peres1995quantum}.

Let's move on to Gleason's Theorem in DQT.
\begin{problem}
Can we recover Gleason's Theorem in DQT? or more precisely: 
\begin{enumerate}
\item Given an assignment $\hat{{\cal P}}:\ffzd{p^{2}}\rightarrow\mathscr{L}_{2}$
in DQT, if $\hat{{\cal P}}$ is a probability, i.e., for any orthonormal
basis $\{\ket{i}\}$, ${\displaystyle \sum_{i=0}^{d-1}\hat{{\cal P}}\left(\ket{i}\right)=1}$,
can we say something about $\hat{{\cal P}}$?
\item Can we induce the Born rule by some reasonable axioms?
\end{enumerate}
\end{problem}
In order to answer the first problem, we need to define the mixed
state in DQT.

\yutsung{The Born rule is the \emph{only}
way to assign a probability in conventional quantum theory because of
Gleason's theorem.  We now show that $\dpr $ defined in
Eq. (\ref{eq:modality}) is an example of a properly defined Born rule
in the discrete theory because $\dpr $ also satisfies the closest
analog of Gleason's Theorem that we can construct.

Following the standard axioms for Gleason's
theorem~\cite{peres1995quantum}, we introduce discrete axioms that
differ in only one significant instance from the classical ones:
\begin{eqnarray}
 & \bullet & \textrm{For any orthogonal basis }\left\{
  \left| n \right\rangle \right\}
  _{n=0}^{d-1}\subset\ffzd{p^{2}}\textrm{, and
  }\label{enu:total-probability-DQT(I)}\\ & &
  \ket{\Psi}\in\ffzd{p^{2}}\textrm{, at least one of\ }
  \dpr \left(\ket{\Psi},\ket{n}\right) = \text{possible.}\nonumber \\ &
  \bullet & \textrm{For any orthogonal basis }
\left\{  \left| n \right\rangle
\right\}_{n=0}^{d-1}\subset\ffzd{p^{2}}\nonumber \\ & &
 \textrm{and a particular\ }\ket{n'}\in\left\{ \left| n \right\rangle \right\}
  _{n=0}^{d-1}\textrm{, }\nonumber \\ & &
  \dpr\big(\ket{n'},\ket{n'}\big)=\text{possible and\ }
   \label{enu:particular-vector-DQT(I)}\\ & &
  \dpr\big(\ket{n'},\ket{n}\big)=\text{impossible for any }n\ne  n'
   \textrm{. }\nonumber \\ & \bullet & \textrm{For any vector  }
   \ket{\Psi},\ket{\Phi}\in\ffzd{p^{2}}
  \textrm{, and  }\gamma\in\ff{p^{2}}\backslash\left\{ 0\right\} 
  \textrm{, }\nonumber
  \\ & &
  \dpr (\ket{\Psi},\ket{\Phi})=\dpr (\gamma\ket{\Psi},\ket{\Phi}) = 
    \dpr (\ket{\Psi},\gamma\ket{\Phi}) \textrm{. }
  \label{enu:parallel-DQT(I)}\\
 & \bullet & \textrm{For any vector
  }\ket{\Psi},\ket{\Phi}\in\ffzd{p^{2}}
  \textrm{, }\nonumber \\ & &
  \textrm{and }U\textrm{ is a }d\textrm{-by-}d
  \textrm{ unitary matrix over }\ff{p^{2}}\textrm{, }
  \label{enu:unitary-DQT(I)}\\ & &
  \textrm{i.e. , }U^{\dagger}U=I\textrm{, then  }
   \dpr (\ket{\Psi},\ket{\Phi})=\dpr (U\ket{\Psi},U\ket{\Phi})\textrm{. }\nonumber
\end{eqnarray}
Because there is no possibility for continuity in the discrete theory,
we have substituted the condition (\ref{enu:unitary-DQT(I)}) for the
conventional continuity axiom in Gleason's Theorem.}

>>>>

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% User specified LaTeX commands.
\usepackage{amsfonts}
\usepackage{color}
\usepackage{stmaryrd}\usepackage{dsfont}
\usepackage[usenames,dvipsnames]{xcolor}
\usepackage{enumitem}		% customizable list environments
\newlength{\lyxlabelwidth}      % auxiliary length 
\theoremstyle{plain}
\newtheorem{thm}{\protect\theoremname}
\ifx\proof\undefined
\newenvironment{proof}[1][\protect\proofname]{\par
\normalfont\topsep6\p@\@plus6\p@\relax
\trivlist
\itemindent\parindent
\item[\hskip\labelsep\scshape #1]\ignorespaces
}{%
\endtrivlist\@endpefalse
}
\providecommand{\proofname}{Proof}
\fi
\theoremstyle{definition}
\newtheorem{problem}[thm]{Problem}

@@

%%%%%%%%%%%%%%%%%
\subsection{\label{sub:A-No-Go-Theorem}A No-Go Theorem}

@@

\yutsung{I am going to prove a no-go theorem for $d=3$ here with
an extra property
\begin{description}
\item [{(iv)}] $\dpt(\ket{\Phi},\ket{\Psi})=\dpt(\gamma\ket{\Phi},\ket{\Psi})=\dpt(\ket{\Phi},\gamma\ket{\Psi})$
where $\gamma\in\ff{p^{2}}^{*}$. 
\end{description}
This property means that the phase difference and normalization will
not effect the probability assignment. Although this property will
not be used explicitly, considering non-normalized states may simplify
the proof. 
\begin{thm}
\label{theorem1-1} There is no map $\dpt\colon\ffzd{p^{2}}\times\ffzd{p^{2}}\longrightarrow[0,1]$
satisfying properties ({\bf i}), ({\bf ii}), ({\bf iii}),
and ({\bf iv}) when $d\ge3$ and $p>3$.\end{thm}
}

@@

%%%%%%%%%%%%%%%%%
\subsection{\label{sub:Interval-Valued-Probabilities-DQT}Interval-Valued
Probabilities for DQT}

@@

Recall the measurement process discussed in section \ref{sec:Critical-Analysis-CQT}.
Given a physical system represented by a state $\ket{\Psi}$, and
an observable ${\mathbf{O}}$, whose eigenspectrum is given by 
\[
{\mathbf{O}}\left|i\right\rangle =\lambda_{i}\left|i\right\rangle \ .
\]
Then the Born rule of CQT tells us that the probability of obtaining
the eigenvalue $\lambda_{i}$ as an outcome of the measurement is
given by 
\begin{eqnarray}
\begin{array}{rcccl}
{\cal P}_{\Psi}\left(\ket{i}\right)={\cal P}_{\Psi}\left(\lambda_{i}\right) & = & {\displaystyle {\frac{\langle\Psi|{P}_{i}|\Psi\rangle}{\ip{\Psi}{\Psi}}}} & = & {\displaystyle {\frac{\ip{\Psi}{i}\ip{i}{\Psi}}{\ip{\Psi}{\Psi}\ip{i}{i}}}}=\Tr\left(\frac{\proj{i}}{\ip{i}{i}}\frac{\proj{\Psi}}{\ip{\Psi}{\Psi}}\right)\end{array}\ ,\label{prob1.eq-6}
\end{eqnarray}
If the system is a mixed state represented by a density matrix 
\[
\rho=\sum_{j}p_{j}\frac{\proj{\Psi_{j}}}{\ip{\Psi_{j}}{\Psi_{j}}}\textrm{ ,}
\]
the Born rule (\ref{prob1.eq-6}) can be extended to 
\begin{eqnarray}
{\cal P}_{\rho}\left(\ket{i}\right)={\cal P}_{\rho}\left(\lambda_{i}\right) & = & \Tr\left(\frac{\proj{i}}{\ip{i}{i}}\sum_{j}p_{j}\frac{\proj{\Psi_{j}}}{\ip{\Psi_{j}}{\Psi_{j}}}\right)=\sum_{i}p_{i}\frac{\ip{\Psi_{j}}{i}\ip{i}{\Psi_{j}}}{\ip{\Psi_{j}}{\Psi_{j}}\ip{i}{i}}\label{prob1.eq-2}
\end{eqnarray}
Let $\C^{d*}=\left\{ \ket{\Psi}\in\C^{d}\mid\ip{\Psi}{\Psi}\ne0\right\} $,
then ${\cal P}_{\rho}:\C^{d*}\rightarrow\left[0,1\right]$ defines
a probability which means ${\cal P}_{\rho}$ satisfies 
\begin{equation}
\textrm{For any orthonormal basis }\{\ket{i}\}\textrm{ and any state }\ket{\Psi}\textrm{ , }{\displaystyle \sum_{i=0}^{d-1}{\cal P}_{\rho}\left(\ket{i}\right)=1}\label{eq:total-probability}
\end{equation}

Conversely, if an assignment ${\cal P}:\C^{d*}\rightarrow\left[0,1\right]$
is a probability, i.e., satisfying equation (\ref{eq:total-probability}),
we can identify ${\cal P}$ to a mixed state $\rho$ by Gleason's
theorem.
\begin{thm}
(Gleason's Theorem) 
\begin{enumerate}
\item For $d\ge3$, for any probability ${\cal P}:\C^{d*}\rightarrow\left[0,1\right]$,
there is a density matrix $\rho$ such that ${\cal P}={\cal P}_{\rho}$~\cite{Redhead1987-REDINA,gleason1957}.\footnote{In order to simplify the statement, we consider consider the probability
on the state vectors instead of the projectors or closed subspaces
as in the Gleason's original paper.}
\item Consequently, the Born rule (\ref{prob1.eq-6}) is the only reasonable
probability assignment, and can be deduced by other reasonable axioms~\cite{peres1995quantum}. 
\end{enumerate}
\end{thm}
Let's move on to Gleason's Theorem in DQT. Can we recover Gleason's
Theorem in DQT? or more precisely: 
\begin{problem}
Given an assignment $\hat{\pi}:\ffzd{p^{2}}\rightarrow\mathscr{L}_{2}$
in DQT, if $\hat{\pi}$ is a probability, i.e., 
\begin{equation}
\textrm{For any orthonormal basis }\{\ket{i}\}\textrm{ , }{\displaystyle \bigvee_{i=0}^{d-1}\hat{\pi}\left(\ket{i}\right)=}\textrm{possible , }\label{eq:total-probability-DQT}
\end{equation}
can we say something about $\hat{\pi}$?
\end{problem}
The answer is no, and the reasons are simple. First, compare the following
arithmetic equation and Boolean equation.
\begin{eqnarray}
p_{1}+p_{2}+p_{3} & = & 1\label{eq:arithmetic}\\
\pi_{1}\vee\pi_{2}\vee\pi_{3} & = & \text{possible}\label{eq:Boolean}
\end{eqnarray}
When $p_{1}=\frac{1}{2}$, we still have an equality $p_{2}+p_{3}=1$
for $p_{2}$ and $p_{3}$ because $\left(\mathbb{R},0,+\right)$ is
a group. In contrast, when $\pi_{1}=\text{possible}$, we know nothing
about $\pi_{2}$ and $\pi_{3}$ because the semigroup$\left(\mathscr{L}_{2},\text{impossible},\vee\right)$
is not cancellative, i.e., $a\vee c=b\vee c$ does not imply $a=b$.
This suggests to study the range of the probability. Consider the
range has the structure $\left(\underline{\mathscr{L}},\mathscr{L},0,1,\oplus,<\right)$,
where $\left(\underline{\mathscr{L}},0,\oplus,<\right)$ is the underlying
commutative ordered semigroup~\cite{clifford1958}, $\mathscr{L}\subseteq\underline{\mathscr{L}}$
is the allowed range value, $1\in\mathscr{L}'\subseteq\mathscr{L}$
is another special value, and for all $x\in\mathscr{L}$, we have
$0<x<1$. For example, $\left(\mathscr{L}_{2},\mathscr{L}_{2},\text{impossible},\text{possible},\vee,<\right)$
and $\left(\mathbb{R},\left[0,1\right],0,1,+,<\right)$ are two possible
range of the probability, but we cannot pick $\underline{\mathscr{L}}=\ff{p^{2}}$
because we cannot define an inequality such that $\left(\ff{p^{2}},0,+,<\right)$
be an ordered semigroup.

Second, Gleason's theorem relies a lot on the continuous probability~\cite{Redhead1987-REDINA,gleason1957,peres1995quantum},
but continuity doesn't even make sense in DQT. This fact implies that
the structure of the domain of the probability, $\ffzd{p^{2}}$, is
also too weak. In order to think which structure we need, let's check
which the reasonable property ${\cal P}_{\Psi}:\C^{d*}\rightarrow\left[0,1\right]$
should have. Despite the total probability is one, ${\cal P}_{\Psi}$
must have 
\begin{eqnarray}
{\cal P}_{\Psi}\left(\ket{i}\right) & = & 0\textrm{ if }\ip{\Psi}{i}=0\label{prob1.eq-6-1}\\
{\cal P}_{\Psi}\left(\ket{i}\right) & = & {\cal P}_{\Psi}\left(\gamma\ket{i}\right)\textrm{ ,}
\end{eqnarray}
where $\gamma\in\C^{*}$. Moreover, the probability must be invariant
under any rotation, i.e., 
\begin{equation}
{\cal P}_{U\Psi}\left(U\ket{i}\right)={\cal P}_{\Psi}\left(\ket{i}\right)\label{eq:rotation}
\end{equation}
for any unitary matrix $U$. Because we consider ${\cal P}_{\Psi}$
and ${\cal P}_{U\Psi}$ together, the first part of Gleason's theorem
doesn't really make sense, but we can still ask whether there is an
unique ``Born rule'' satisfying all the conditions. If we add these
three conditions, our problem becomes
\begin{problem}
\label{prob:Gleason's}Is there a structure $\left(\underline{\mathscr{L}},\mathscr{L},0,1,\oplus,<\right)$
such that $\tilde{\pi}:\ffzd{p^{2}}\times\ffzd{p^{2}}\rightarrow\mathscr{L}$
satisfying the following conditions is unique for $d\ge3$ and infinitely
many for $d=2$? 
\begin{eqnarray}
 & \bullet & \textrm{For any orthonormal basis }\left\{ \ket{i}\right\} _{n=0}^{d-1}\subset\ffzd{p^{2}}\textrm{, }\label{enu:total-probability-DQT(I)}\\
 &  & \textrm{ \ensuremath{{\displaystyle \bigoplus_{i=0}^{d-1}\tilde{\pi}\left(\ket{\Psi},\ket{i}\right)=1}} .}\nonumber \\
 & \bullet & \textrm{If }\braket{\Psi}{i}=0\textrm{ ,}\nonumber \\
 &  & \textrm{then }\tilde{\pi}\left(\ket{\Psi},\ket{i}\right)=0\textrm{ .}\label{enu:particular-vector-DQT(I)}\\
 & \bullet & \textrm{For any vector }\ket{\Psi},\ket{i}\in\ffzd{p^{2}}\textrm{, }\nonumber \\
 &  & \textrm{and }U\textrm{ is a }d\textrm{-by-}d\textrm{ unitary matrix over }\ff{p^{2}}\textrm{, }\label{enu:unitary-DQT(I)}\\
 &  & \textrm{i.e. , }U^{\dagger}U=I\textrm{, then }\tilde{\pi}(\ket{\Psi},\ket{i})=\tilde{\pi}(U\ket{\Psi},U\ket{i})\textrm{. }\nonumber \\
 & \bullet & \textrm{For any vector }\ket{\Psi},\ket{i}\in\ffzd{p^{2}}\textrm{, and }\gamma\in\ff{p^{2}}^{*}\textrm{, }\nonumber \\
 &  & \dpr(\ket{\Psi},\ket{\Phi})=\dpr(\gamma\ket{\Psi},\ket{\Phi})=\dpr(\ket{\Psi},\gamma\ket{\Phi})\textrm{. }\label{enu:parallel-DQT(I)}
\end{eqnarray}

\end{problem}
The no-go theorem in section \ref{sub:A-No-Go-Theorem} said the if
$\left(\underline{\mathscr{L}},\mathscr{L},0,1,\oplus,<\right)=\left(\mathbb{R},\left[0,1\right],0,1,+,<\right)$,
there is no $\tilde{\pi}:\ffzd{p^{2}}\times\ffzd{p^{2}}\rightarrow\mathscr{L}$
for $d\ge3$. It is easy to extend the no-go theorem and prove 
\begin{thm}
If $\left(\underline{\mathscr{L}},0,\oplus,<\right)$ is a ordered
group, then there is no $\tilde{\pi}:\ffzd{p^{2}}\times\ffzd{p^{2}}\rightarrow\mathscr{L}$
satisfying all conditions in problem \ref{prob:Gleason's} for $d\ge3$.
\end{thm}
When we consider $\left(\mathscr{L}_{2},\mathscr{L}_{2},\text{impossible},\text{possible},\vee,<\right)$,
$\dpr$ defined in section \ref{sub:Interval-Valued-Probabilities-DQT}
satisfies all conditions, but $\dpr$ is not the unique solution because
of lack of cancellation law. It is still an open question whether
problem \ref{prob:Gleason's} is true or false in general.
